### 2024-04-03 Chapters/Chapter1.tex
Uppfært eftir ábendingar frá ÓPP

%----------------------------------------------------------------------------------------
Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed. The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period. The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute interval. This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}. As the Navier Stokes Equation (\ref{eq:navierstokes}) shows the variability of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth, and resistance. It states that the change in wind is described by the pressure gradient, the oscillation, the acceleration due to gravity and the resistance from the landscape.

\begin{equation}
    \label{eq:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
\end{equation}

Equation \ref{eq:navierstokes} relates the change in wind w.r.t. time ($\frac{\delta \boldsymbol{V}}{\delta t}$) and change in wind w.r.t. position ($\nabla\boldsymbol{V}$) to pressure change w.r.t. position ($\nabla P$). The wind is also influenced by the oscillation, gravity and resistance but the main drive of wind is pressure gradient\cite{uncertainties_in_numerical_weather_predictions}.

@@

%This thesis looks at how best to predict gust factor based on wind speed, wind direction, pressure and temperature at different heights above ground, as well as the elevation of the landscape over an area in the direction the wind is coming from. A neural network (NN) was constructed to predict the gust factor.

@@

\begin{enumerate}[label = Rule \arabic*:]
    \item Large Parameter Count. The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
    \item Large Number of Predictands: predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
    \item Scalability and downstream applicability. This might crystallize in predicting cyclones. Often, the teams responsible for creating these models try to show their applicability to predict cyclones when not trained specifically on cyclone data (e.g.  \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast})\cite{SecondRevolution}. This is done to show the versability of the models.
\end{enumerate}

Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training. No model had shown that it could in any way completely replace the traditional systems. In early 2022, Pathak et al. presented FourCastNet. FourCastNet uses an Adaptive Fourier Neural Operator model that leverages transfomer architecture rather than the popular convolutional model architechture. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms for smaller variables. It generates a week-long forecast in less than 2 seconds, orders of magnitude faster than standard physical methods\cite{FourCastNet}. In 2022, machine learning methods were presented that made predictions much faster than traditional NWP, after a one time training (or at least training that wouldn't have to be redone often). These were in some cases performing better than NWP.  In 2023, Remi Lam and the GraphCast team at Google introduced GraphCast. This model was able to outperform the industry standard High Resolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This model as the name suggests leverages graphing connections rather than traditional grid like data structure. The base data is given in latitude and longitude degrees at a resolution of 0.25 degrees. This means points are closer to each other at the poles. Using the graphing structure is supposed to help with bias incurred as a result of this\cite{GraphCast}.

@@

\section{Methodology and related work}

This study looks at data from three sources and an attempt is made to predict the gust factor in a given place in Iceland. It uses reanalysis data, along with elevation data to predict the gust factor. It looks at the data at the point of interest and does not look at the data as a time series. This thesis aims to improve on the baseline model of always predicting the mean gust factor and show that some structure can be learned from reanalysis data about gusts. To do this a neural network was created. Any significant improvement on a base model, that always guesses the mean gust factor, would indicate that the final model has something to contribute.

@@

To be able to capture the patterns in the data a neural network was constructed. A NN architechture was chosen as they are known to be able to capture patterns well in complex data and handle high parameter counts. This comes in handy when training on different types of data. It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset. To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE, \ref{eq:mape})was used. 
\begin{equation}
    \label{eq:mape}
    MAPE = \frac{}{}
\end{equation}

This was chosen because the target is the gust factor (the wind gust over the average wind). If the target would have been the wind gust rather than the gust factor then something like mean absolute error might be more appropriate.

>>>>

%----------------------------------------------------------------------------------------
Wind gusts are brief increase in wind speed (lasting seconds) as compared to mean wind speed. The gust factor is defined as the peak gust divided by the mean wind speed over some defined time period. The peak wind gust is often defined as the highest 3 second rolling average measured wind speed over a period of 10 minutes, while the mean wind is the average of all measurements in the 10 minute interval. This varies, with the US using a 1 minute interval, leading to 14\% higher results \cite{why_wind_gusts}. As the Navier Stokes Equation (\ref{eq:navierstokes}) shows the variability of the wind, in time and space, is dependent upon the pressure gradient, the oscillating force of the earth, and resistance. It states that the change in wind is described by the pressure gradient, the oscillation, the acceleration due to gravity and the resistance from the landscape.
\begin{equation}
    \label{eq:navierstokes}
    \frac{\delta \mathbf{V}}{\delta t} + \mathbf{V}\cdot\nabla\mathbf{V} = \underbrace{-\frac{1}{\rho}\nabla P}_{pressure} -\overbrace{ f\mathbf{k}x\mathbf{V}}^{oscillation} - g - \underbrace{\frac{\delta(u'\omega')}{\delta z} - \frac{\delta(v'\omega')}{\delta z}}_{resistance}
\end{equation}
Equation \ref{eq:navierstokes} relates the change in wind w.r.t. time ($\frac{\delta \boldsymbol{V}}{\delta t}$) and change in wind w.r.t. position ($\nabla\boldsymbol{V}$) to pressure change w.r.t. position ($\nabla\boldsymbol{P}$). The wind is also influenced by the oscillation, gravity and resistance but the main drive of wind is pressure gradient\cite{uncertainties_in_numerical_weather_predictions}.

@@

@@

\begin{enumerate}[label = Rule \arabic*:]
    \item Large Parameter Count. The number of parameters can vary wildly but a general range might be from tens of millions to billions of parameters
    \item Large Number of Predictands: predicting on different levels (such as pressure levels or height levels) and offering detailed information on the atmospheric vertical structure and surface conditions
    \item Scalability and downstream applicability. This might crystallize in predicting cyclones. Often, the teams responsible for creating these models try to show their applicability to predict cyclones when not trained specifically on cyclone data (e.g.  \href{https://www.youtube.com/watch?v=PD1v5PCJs_o&ab_channel=GregBronevetsky}{GraphCast})\cite{SecondRevolution}. This is done to show the versability of the models.
\end{enumerate}
Before 2022, LWM had been shown to be able to compete with traditional NWP for some specifc cases as well as making predictions quicker, after training. No model had shown that it could in any way completely replace the traditional systems. In early 2022, Pathak et al. presented FourCastNet. FourCastNet uses an Adaptive Fourier Neural Operator model that leverages transfomer architecture rather than the popular convolutional model architechture. FourCastNet matches the performance of standard forecasting techniques at short lead times for large-scale variables and outperforms for smaller variables. It generates a week-long forecast in less than 2 seconds, orders of magnitude faster than standard physical methods\cite{FourCastNet}. In 2022, machine learning methods were presented that made predictions much faster than traditional NWP, after a one time training (or at least training that wouldn't have to be redone often). These were in some cases performing better than NWP.  In 2023, Remi Lam and the GraphCast team at Google introduced GraphCast. This model was able to outperform the industry standard High Resolution Forecast (HRES) produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This model as the name suggests leverages graphing connections rather than traditional grid like data structure. The base data is given in latitude and longitude degrees at a resolution of 0.25 degrees. This means points are closer to each other at the poles. Using the graphing structure is supposed to help with bias incurred as a result of this\cite{GraphCast}.

@@

\section{Methodology and related work}
This study looks at data from three sources and an attempt is made to predict the gust factor in a given place in Iceland. It uses reanalysis data, along with elevation data to predict the gust factor. It looks at the data at the point of interest and does not look at the data as a time series. This thesis aims to improve on the baseline model of always predicting the mean gust factor and show that some structure can be learned from reanalysis data about gusts. To do this a neural network was created. Any significant improvement on a base model, that always guesses the mean gust factor, would indicate that the final model has something to contribute.

@@

To be able to capture the patterns in the data a neural network was constructed. A NN architechture was chosen as they are known to be able to capture patterns well in complex data and handle high parameter counts. This comes in handy when training on different types of data. It is also easy to construct different types of neural networks and see how they fit well with parts of the dataset. To measure the performance of these models, both to train and test, mean absolute percentage error (MAPE) as defined in Equation (\ref{eq:mape}) was used.
\begin{equation}
    \label{eq:mape}
    MAPE = \frac{1}{n}\Sigma_{i=1}^n\frac{|y_{predict} - y_{true}|}{y_{predict}}
\end{equation}
This was chosen because the target is the gust factor (the wind gust over the average wind). If the target would have been the wind gust rather than the gust factor then something like mean absolute error might be more appropriate.

### 2024-04-03 Chapters/Chapter2.tex
Uppfært eftir ábendingar frá ÓPP

\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

@@

Veðurstofan provided recordings from weather stations all around Iceland. These automatic weather stations (AWS) are at a height of 10 meters above ground level. The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents. The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These features for each of the stations go back at least two decades, but do not have the same start date. This paper does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\section{Carra Data}
The CARRA dataset goes back to September 1990 and is currently updated monthly, with a latency of 2-3 months. The oldest Vedurstofu data example that fulfills given criteria is from \startDateVedur. This is covered by Carra. A few of the newer points from AWS were not available from Carra when data was gathered. The Carra dataset is available for two regions, west and east. Each of these covers a vastly larger area than the area of interest. It is possible to request a subsection of the area, a rectangular subsection of the area. This leads to having to store a large amount of data, because we cannot simply ask for the specific points of interest. To get the data one has two options. Their web interface or using their API client. Using the API client is the only realistic option here, as we are making thousands of requests for different times.\\

Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit 
m/s over a 10 minute interval). For each such observation in the Veðurstofa gögn, we need to make two API (calls excluding redundancy). One for the 3 hour interval before and one for the 3 hour interval after. That is if the observation was at 1 pm, we would need Carra data for noon and 3 pm. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, we should be able to bridge and get an estimation for the point of the given weather station. The Carra data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground.\\

For each point 4 features were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be bridged to create data for model to be trained on.

After using this method to request terabytes of data, it was discovered that it is apparently possible to query a specific area. This decreased the size of each file created after a request from around 50 MB to around 2 MB. Downloading and processing this data would take a lot less time. The largest bottleneck in terms of retrieving the data is the request time. That is the time the request is queued and the time it is running before it begins downloading. This could range from almost immediate to 15 minutes or more. This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring, which increased the wait time and sometimes resulted in queries not being responded to. This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project.

\section{Elevation data}
Veðurstofa provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid. This file encompasses Iceland and is around 685 Mb. At first I thought I might have to split the file into subsections and read for each one but that doesn't seem to be the necessary. A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area. The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes. A quadtree was constructed for each of the sections, so as to simplify the lookup. A quadtree allows for efficient lookup by searching for points that intersect a given boundary. A quadtree, like the name indicates, is a tree structure. To initialize it you give it a boundary box. This boundary box is then the parent of exactly four nodes and so on, dividing the area in four at each level and allowing quick lookup. This was not necessary as the rasterio package allowed for quick lookup with it's index and the affine transform.

>>>>

\label{Chapter2} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

@@

Veðurstofan provided recordings from weather stations all around Iceland. These automatic weather stations (AWS) are at a height of 10 meters above ground level. The information that is provided by these AWS is presented in two different type of documents, hourly and 10 minute documents. The hourly documents are summations of the 10 minute documents, with the exception that any errors (so called nails) still in the 10 minute documents should have been removed from the hourly documents. Each type of document contain the following information: the date and time, the station number (that can be converted to the coordinates using another document), the average wind speed, the wind gust, the standard deviation of the wind gust, the direction of the wind and the standard deviation for the wind direction. These features for each of the stations go back at least two decades, but do not have the same start date. This thesis does not look at the data as a time series, it tries to make predictions using only the information at a given point in time.

\section{Carra Data}
The CARRA dataset goes back to September 1990 and is currently updated monthly, with a latency of 2-3 months. The oldest Vedurstofu data example that fulfills given criteria is from \startDateVedur. This is covered by Carra. A few of the newer points from AWS were not available from Carra when data was gathered. The Carra dataset is available for two regions, west and east. Each of these covers a vastly larger area than the area of interest. It is possible to request a subsection of the area, a rectangular subsection of the area. This leads to having to store a large amount of data, because it is not possible to ask for specific points of interest. To get the data one has two options. Their web interface or using their API client. Using the API client is the only realistic option here, as there were thousands of requests made for different times.

Each request is made by going through all instances of the Veðurstofu data were able to fulfill the requirements (of mean wind speed reaching \averageWindSpeedLimit m/s over a 10 minute interval). For each such observation in the Veðurstofa data, two API calls excluding redundancy are needed. One for the 3 hour interval before and one for the 3 hour interval after, if the data point doesn't fall exactly one of the three hour interval times. That is if the observation was at 13:00, we would need Carra data for noon and 15:00. Date and times were generated automatically from the AWS and requested by calling the client. Using these datapoints, interpolation was used to get an estimation for the point of the given weather station. The Carra data contains several types of layers. These are single levels, model levels, height levels, pressure levels. The data for this observation was downloaded from height levels. That is, data was requested at heights of 15, 150, 250 and 500 meters above ground. For each point 4 parameters were requested, wind speed, wind direction, pressure and temperature. Each of these features needed to be interpolated to create data for model to be trained on.

After using this method to request terabytes of data, it was discovered that it is apparently possible to query a specific area. This decreased the size of each file created after a request from around 50 MB to around 2 MB. Downloading and processing this data would take a lot less time. The bottleneck for retrieving the data is the request time. That is the time the request is queued and running (server preparing the data) before it begins downloading. This could range from almost immediate to 30 minutes. This problem was exacerbated by the fact that the climate data store (CDS) was undergoing updates during the winter and spring of 2023/2024, which increased the wait time and sometimes resulted in queries not being responded to. This meant that the time it would take to retrieve the remaining information went from around a day, when the requests were at their quickest, to many months, something that would not be possible given the time frame of the project. Fortunately these problems were only particularly time consuming during mid winter.

\section{Elevation data}
Veðurstofa provided a tif file containing the elevation of Iceland on a 20 meter by 20 meter grid. This file encompasses Iceland and is around 685 Mb. A good amount of time was spent finding a data structure that was best for lookup when trying to find points within a certain area. The country was divided into 10 parts (as only around 13\% of the file was able to be read into memory at each time as a part of dictionary object) with boundary boxes. A quadtree was constructed for each of the sections, so as to simplify the lookup. A quadtree allows for efficient lookup by searching for points that intersect a given boundary. A quadtree, like the name indicates, is a tree structure. To initialize it a boundary box is given. This boundary box is then the parent of exactly four nodes and so on, dividing the area in four at each level and allowing quick lookup. This was not necessary as the Python package Rasterio allowed for quick lookup with it's index and the affine transform. Using this package it is possible to quickly look up elevation given coordinates using matrix calculations.

### 2024-04-03 Chapters/Chapter3.tex
Uppfært eftir ábendingar frá ÓPP

This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models. When working with hundred of thousands of rows, the efficiency of the code is very important. Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible. The three data sources were all in a different format. Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from Carra was in a GRIB format. To use the data to train, these three data sources needed to be combined into one file. This was done based on the measurement data from the Veðurstofa. A limit was set on the average wind speed and it was used to select measurement points. Along with the average wind speed having to be above a certain limit, to make sure that we were not essentially getting duplicates, we would select only the top wind speed in any given 48 hour period. That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times. The data from Veðurstofan was supplied for 10 minute increments, while Carra data is in 3 hour intervals. This means that to use the Carra data to predict the measured values from Veðurstofa, temporal bridging would need to be done. Along with the temporal bridging, we note that the Carra data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations. The elevation information was given by a 20 by 20 rectangular grid that covers Iceland. When combining these data sources a selection of bridging needs to occur. The bridging was linear, both temporally and spatially. This might influence the results but was not considered in this study.

The procedure of combining these sources was as follows. We start with the measured data from the AWS, we filter this by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on the factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there we have the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API. When querying the Carra API we will have to query in such a way that we query for given hours, days, months, years and a given area. That is, if we ask for a given hour, we get that hour for every day that we ask for. Similarly if we ask for a given day, we get that day for every month we ask for. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for bridging if close to midnight) but every hour of the day (midnight, 3 AM, 6 AM, 9 AM, midday, 3 PM, 6 PM and 9 PM) and then take these values and bridge for the 10 minute values. After querying and downloading the data for the height levels and variables requested, bridging is done for the points of interest and values stored in a pandas dataframe. After this is done the downloaded data is discarded and we look to the next month. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

The elevation data comes in a GeoTif file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved bridging needs to be done. This is done in a similar manner to the briding of the Carra data. Weights are assigned to each of the four points bounding the point of interest. These weights are then used to calculate the weighted average that represents the bridged value. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

@@

Once data has been retrieved for all three sources and processed, including bridging values, it needs to be made ready to use by the model, for both training, validation and test. We start with a dataframe that contains the measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the Carra data we choose certain height levels. That is at which heights we want the reanalysis data. These present as separate lines in the Carra dataframe. We need to combine them so that each line indicates a single observation. When this is done we can combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns. Lastly we look at the elevation data. A couple of different sections of land around the weather stations have been looked at. We tried a sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, was selected like as shown in code listing \ref{code:sectorElevation}. That is we look at the wind direction d and define a range of angles around that direction at some distance from the given point. This means that we get some number of points (equal to the length of angleRange) at a distance from given weather station as defined by length\_rng. This then gives us some sectors at some distances away from the weather station.

@@

We end up with a dataframe that has measured data from AWS, which gives us our target, reanalysis data from Carra, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data. An example of what the data looks like can be seen in table \ref{table:trainDataExample}.

\begin{table}[h]
    \resizebox{\textwidth}{!}{
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
         Ri\_01 & Ri\_12 & N\_01 & N\_12 & station\_elevation & relative\_corner & PC1 & ...\\\hline
         -1.18e+00 &  2.67e+04 & -8.57e-06 & 6.78e-05 & 3.34e+01 & 2.73e+00 & 1.36e+01 & ... 
    \end{tabular}
    }
    \caption{An example of data structure used with model}
    \label{table:trainDataExample}
\end{table}

Looking at table \ref{table:trainDataExample} we note that the last \nPCA columns represent the principle component analysis of the elevation data. We also note that the first four columns represent two variables that describe the stability of the air. These are the Richardson number (Ri) and Brunt–Väisälä frequency (N). They are calculated using equations \ref{eqn:Ri} and \ref{eqn:N}. They are calculated using reanalysis data about the the weather at two different height levels. Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1. Exactly the same notation is used with the Brunt–Väisälä frequency. The height levels used were 15, 250 and 500 meters above the ground.

\begin{equation}
    \label{eqn:Ri}
    Ri = \frac{g \cdot dT \cdot dz}{T_{ave} \cdot dU^2}
\end{equation}

\begin{equation}
    \label{eqn:N}
    N = \sqrt{\frac{g \cdot dT }{T_{ave} \cdot dz}}
\end{equation}

Here, g is the acceleration due to gravity, dT is the temperature difference between the two height levels, dz is the elevation difference, $T_{ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and dU is the wind speed difference between the two height levels. Both of these numbers tell us something about the stability of the air. These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain in calculating the as opposed to having the raw data. Including these factors instead of the every variable might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability.

>>>>

This project used three main data sources, which need to be queried, filtered and combined to prepare the data for use in the models. When working with hundred of thousands of rows, the efficiency of the code is very important. Iterating through those rows might be necessary at times but will increase the time exponentially as compared to using vectorizing methods were possible. The three data sources were all in a different format. Measurement data from Veðurstofa was in text files, elevation data was in GeoTiff and reanalysis data from Carra was in a GRIB format. To use the data to train, these three data sources needed to be combined into one file. This was done based on the measurement data from the Veðurstofa. A limit was set on the average wind speed and it was used to select measurement points. Along with the average wind speed having to be above a certain limit, to make sure that we were not essentially getting duplicates, we would select only the top wind speed in any given 48 hour period. That is, for a data point to be included it must have been the highest wind speed in the previous and following 24 hours, so as to not describe the same weather multiple times. The data from Veðurstofan was supplied for 10 minute increments, while Carra data is in 3 hour intervals. This means that to use the Carra data to predict the measured values from Veðurstofa, temporal interpolation would need to be done. Along with the temporal interpolation, note that the Carra data is given in a rectangular grid where the distance between each point is around 2.5km while the the information from the Veðurstofa is given at specific locations. The elevation information was given by a 20 by 20 rectangular grid that covers Iceland. When combining these data sources a selection of interpolation method needs to occur. The interpolation was linear, both temporally and spatially. This might influence the results but was not considered in this study.

The procedure of combining these sources was as follows. The measured data from the AWS is filtered by using a limit on the average wind speed. The gust factor generally drops with increased wind speed (although not always dependent on factors such as the landscape \cite{GNP_vidtal}). Even so being able to predict the gust factor is more important for higher average wind speed as there we have the highest wind gusts. After this stripped dataset over every AWS has been created it is used to query the Carra data by using their API. The Carra API needs to be queried for given hours, days, months, years and a given area. That is, if queried for a given hour, it returns that hour for every day that is queried. Similarly if queried for a given day, it returns that day for every month. In light of these restraints, it was decided to query month by month. Querying only the days needed (both the days included in the 10 minute measurements and the days needed for bridging if close to midnight) but every hour of the day (midnight, 03:00, 06:00, 09:00, midday, 15:00, 18:00 and 21:00 PM) and then take these values and bridge for the 10 minute values. After querying and downloading the data for the height levels and variables requested, bridging is done for the points of interest and values stored in a pandas dataframe. After this is done the downloaded data is discarded and the next month is queried. This drastically decreases the amount of data that needs to be stored as compared to downloading the entire area and keeping all the data points (goes from several terabytes to less than a gigabyte).

The elevation data comes in a GeoTif file that covers Iceland. It is a rectangular grid of resolution 20 meters. For every point of interest (every weather station), the elevation of that given point along with other points surrounding the weather station is retrieved. For each point retrieved bridging needs to be done. This is done in a similar manner to the interpolation of the Carra data. Weights are assigned to each of the four points bounding the point of interest. These weights are then used to calculate the weighted average that represents the bridged value. This information is included in the training data as the landscape is known to influence both the average wind and the gustiness \cite{GNP_vidtal}.

@@

Once data has been retrieved for all three sources and processed, including bridging values, it needs to be made ready to use by the model, for both training, validation and test. Starting with a dataframe that contains measured information from AWS. This includes the average wind, the wind gust, wind direction along with the station number and coordinates. When selecting the Carra data certain height levels are chosen. That is at which heights above ground we want the reanalysis parameters that are requested. These present as separate lines in the Carra dataframe. We need to combine them so that each line indicates a single observation. When this is done it is possible to combine the AWS Veðurstofu data and Carra reanalysis data on the location and time columns. The last data source is the elevation. A couple of different sections of land around the weather stations have been looked at. A sector of a circle looking upwind, two sectors looking upwind and downwind and a circle around the point. In any case the points, that represent these sections, were selected like as shown in code Listing (\ref{code:sectorElevation}). That is a range of angles are defined based on the wind direction d at some distance from the given point. This means that the resultant points (equal in number to the length of angleRange by k) form sectors at several distances from the given weather station.

@@

We end up with a dataframe that has measured data from AWS, which gives us our target, reanalysis data from Carra, which gives us weather variables to train on, and finally elevation points in the landscape to include in our training data. An example of what the data looks like can be seen in Table (\ref{table:trainDataExample}).

\begin{table}[h]
    \caption{An example of data structure used with model}
    \label{table:trainDataExample}
    \resizebox{\textwidth}{!}{
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
         Ri\_01 & Ri\_12 & N\_01 & N\_12 & station\_elevation & relative\_corner & PC1 & ...\\\hline
         -1.18e+00 &  2.67e+04 & -8.57e-06 & 6.78e-05 & 3.34e+01 & 2.73e+00 & 1.36e+01 & ... 
    \end{tabular}
    }
\end{table}

Looking at Table (\ref{table:trainDataExample}) note that the last \nPCA columns represent the principle component analysis of the elevation data. Also note that the first four columns represent two variables that describe the stability of the air. These are the Richardson number ($Ri$)\cite{richardson_number_skybrary} and Brunt–Väisälä frequency\cite{brunt_vaisala_freq_eumtrain} ($N$). They are calculated using Equations (\ref{eqn:Ri}) and (\ref{eqn:N})\cite{mean_gust_HA_HO}. They are calculated using reanalysis data about the the weather at two different height levels. Thus $Ri_{01}$ refers to the Richardson number calculated between height levels 0 and 1. Exactly the same notation is used with the Brunt–Väisälä frequency. The height levels used were 15, 250 and 500 meters above the ground.

\begin{equation}
    \label{eqn:Ri}
    Ri = \frac{g \cdot dT \cdot dz}{T_{\textrm ave} \cdot dU^2} \unit{[]}
\end{equation}

\begin{equation}
    \label{eqn:N}
    N = \sqrt{\frac{g \cdot dT }{T_{\textrm ave} \cdot dz}} \unit{[Hz]}
\end{equation}

Here, $g$ is the acceleration due to gravity, $dT$ is the temperature difference between the two height levels, dz is the elevation difference, $T_{\textrm ave}$ is the average temperature (that is the average of the two temperatures in the height levels) and $dU$ is the wind speed difference between the two height levels. Both of these numbers provide some insight about the stability of the air. A lower value for the Richardson number indicates a higher turbulance. A typical range of values could be between 10 and 0.1, with values below 1 indicating significant turbulance\cite{richardson_number_skybrary}. When the square of the Brunt-Väisälä frequency is negative, then we have unstable air (air parcel will move away from its original position)\cite{brunt_vaisala_freq_eumtrain}. These are derived factors from the reanalysis data and as such there shouldn't be a significant information gain using $Ri$ and $N$ as opposed to having the raw data. Including these factors instead of every reanalysis variable requested might speed up training as well as making the model more easily explainable with the use of Shapley values or other tools for explainability. Using Shapley a feature importance value is attributed to a given feature by creating all possible permutations of any possible length (up to number of features) and seeing how the predictions are skewed when the given parameter is included or excluded. This the needs to be done for all parameters. The time complexity of this is very high ($2^n$ coalitions)\cite{shapley_information}. Most implementations use some approximations, which still can take a considerable amount of time for models with a high parameter count and many examples.
